### 2022-10-13 AISTATS2023/paper.tex
latest

We perform a two-fold evaluation of UALE. The first group
(Section~\ref{sec:simulation-examples-1}) compares UALE with PDP-ICE,
the main feature effect method that also quantifies the heterogeneity
of instance-based effects. The second group
(Section~\ref{sec:simulation-examples-2}) evaluates UALE
approximation, i.e., compares UALE's optimal bin-splitting against the
fixed-size alternative.

\subsection{Case 1: UALE vs PDP-ICE}
\label{sec:simulation-examples-1}

This a qualitative example with the aim to highlight that PDP-ICE is
vulnerable to misleading estimations of both the average effect and
the uncertainty, even when the black-box function is relatatively
simple.

We use the data generating distribution
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
linear function

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_1(\mathbf{x}) < 1$}\\
                    \alpha f_2(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The part \(f_1\) is a linear function of
the two correlated features, \(x_1, x_2\), and \(f_2\) is an
interaction term between the two non-correlated, \(x_1, x_3\).  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) with
equal weights (\(a_1=a_2\)).

In the general case, there is no unique ground-truth feature effect
and uncertainty and each method defines the effect in a different
way. We select this simple case, to determine the ground-truth,
exploiting two facts. First, in \(p(\xb)\) it holds that
\(x_1=x_2\). Therefore, knowing the value of either \(x_1\) or
\(x_2\), we know the closed-form of \(f(\xb)\), too. For example, if
\(a_1=a_2=1\) and \(x_1 = \frac{1}{6}\), then
\(f_1(\xb) = \frac{1}{3}\), so
\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \). Afterwards, we use the
properties of interpretable models~\citep{molnar2022} to determine the
ground-truth. In the example above, if also \(a=0\), then
\(f(\xb) = x_1 + x_2\) which is an interpretable linear model. For the
uncertainty, we use the fact that under no-interactions, \(a=0\), the
uncertainty must be zero (no heteregoneity). We discuss separately the
case with \(a>0\). Finally, we want to also show that UALE's
approximation is accurate, due to appropriate bin splitting.

\paragraph{No Interaction, Equal weights.}

In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
(equal weights). We examine only the effect of feature \(x_1\),
because it is exactly the same with \(x_2\). Based on the discussion
above, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is:
\(x_1\) in \([0, \frac{1}{4}]\), \(-x_1\) in
\([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in \([\frac{1}{2},
1]\). Because \(x_1\) does not interact with any other variable, the
uncertainty is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in
\([0,1]\). In Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP
effect is wrong and ICE plots show heterogeneous effects. In contrast,
UALE models correctly both the average effect and the absence of
uncertainty. Furthermore, we observe that bin-splitting is accurate
(zero cumulative error \(\mathcal{E}_{\mathcal{Z}}\)).

@@

\paragraph{No Interaction, Non-Equal Weights.}

In this set-up, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= 0.5\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, 0.2]\), \(-2x_1\) in \([0.2, 0.4]\) and \(0\) in \([0.4,
1]\). As before, the ground-truth uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is opposite to the ground-truth effect,
i.e.~negative in the region \([0, 0.2)\), positive in \([0.2, 0.4)\),
and the ICE erroneously implies the existence of heterogeneous
effects. As before, ALE quantifies correctly the ground truth effect,
the zero-uncertainty and partioning correctly the \(x_1\) domain.

@@

\paragraph{With Interaction, Equal weights.}

In this set-up, we activate the interaction term, i.e.  \(a=1\),
keeping the weights equal \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects for features \(x_1\) and \(x_3\), because the
local effects at \(x_1\) depend on the unknown value of \(x_3\) and
vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
not appear in any interaction term.  As we observe in
Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
and the uncertainty in all cases.  On the other hand, PDP-ICE
quantifies correctly the average effect and the uncertainty only in
the case of feature \(x_3\), because it is independent from other
features. For the correlated features \(x_1\) and \(x_2\) both the
average effect (PDP) and the uncertainty (ICE) are wrong.

@@

Despite the model's simplicity, PDP-ICE fail in modeling both the
average effect and the uncertainty. The error is due to the method's
formulation, i.e.~the use of the marginal distribution that ignores
correlations between features, and not due to poor approximation
because of limited samples. In contrast, UALE models both correctly
and estimates them accurately using an optimal bin splitting in terms
of Eq.~(\ref{eq:opt}). The examples above do not cover the case of an interaction term
between correlated features, for example a term \(x_1x_2\), because
there is an open debate about the ground-truth effect in this
case~\citep{Gromping2020MAEP}.

>>>>

We perform a two-fold evaluation of UALE. First,
(Section~\ref{sec:simulation-examples-1}) compares UALE with PDP-ICE,
the main feature effect method that also quantifies the heterogeneity
of instance-level effects. Second,
(Section~\ref{sec:simulation-examples-2}) evaluates UALE
approximation, i.e., compares UALE's optimal bin-splitting against the
fixed-size alternative.

\subsection{UALE vs PDP-ICE}
\label{sec:simulation-examples-1}

This a qualitative example to highlight that PDP-ICE is vulnerable to
misleading estimations of both the average effect and the uncertainty,
even in case of a simple black-box function, as opposed to UALE.

\paragraph{Example set-up.}
We use the data generating distribution
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\), where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\), and the following function,

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_1(\mathbf{x}) < 1$}\\
                    \alpha f_2(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The part \(f_1\) is a linear function of
the two correlated features, \(x_1, x_2\), and \(f_2\) is an
interaction term between the two non-correlated, \(x_1, x_3\).  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) and
equal weights (\(a_1=a_2\)).

In the general case, evaluating feature effect methods is a
challenging task, because there is no unique ground-truth feature
effect and uncertainty and each method defines the effect in a
different way. To deal with this, we exploit two characteristics of
the above set-up. First, in \(p(\xb)\) it holds that
\(x_1=x_2\). Therefore, knowing \(x_1\) or \(x_2\), we know the
closed-form of \(f(\xb)\), too. For example, if \(a_1=a_2=1\) and
\( 0 \leq x_1 < \frac{1}{4}\), then \(f_1(\xb) < \frac{1}{2}\), so
\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \), i.e., the first branch of
Eq.~\ref{eq:synth-ex-1-function}. Afterwards, we can determine the
ground-truth. In the example above, if \(a=0\), then
\(f(\xb) = x_1 + x_2\) and therefore the ground-truth effect of
\(x_1\) is \(f^{\mathcal{GT}}(x_1) = x_1\) in the corresponding
interval, i.e. \(x_1 \in [0, \frac{1}{4})\). For the uncertainty, we
use the fact that under no-interactions, \(a=0\), the uncertainty must
be zero (no heteregoneity). We discuss separately the case with
\(a>0\). Finally, we also demonstrate that UALE's approximation is
accurate, due to the bin splitting that we propose.

\paragraph{(a) No Interaction, Equal weights.}

Here, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\) (equal
weights). We examine only the effect of feature \(x_1\), because it is
exactly the same as with \(x_2\). Based on the discussion above, the
ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) in
\([0, \frac{1}{4}]\), \(-x_1\) in \([\frac{1}{4}, \frac{1}{2}]\) and
\(0\) in \([\frac{1}{2}, 1]\). Because \(x_1\) does not interact with
any other feature, the uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots show heterogeneous effects. In contrast, UALE
models correctly both the average effect and the absence of
uncertainty. Finally, we observe that UALE's bin-splitting leads (a)
to an accurate uncertainty estimation (near-zero cumulative bin-error
\(\mathcal{E}_{\mathcal{Z}}\)) and (b) an easier interpretation,
creating three wide bins, i.e.
\([0, \frac{1}{4}), [\frac{1}{4}, \frac{1}{2}), [\frac{1}{2}, 1)\),
that correspond to the piecewise-linear regions of the effect.

@@

\paragraph{(b) No Interaction, Non-Equal Weights.}

Here, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= \frac{1}{2}\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, \frac{1}{5}]\), \(-2x_1\) in \([\frac{1}{5}, \frac{2}{5}]\) and
\(0\) in \([\frac{2}{5}, 1]\). As before, the ground-truth uncertainty
is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with any other feature. In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is opposite to the ground-truth effect,
i.e.~negative in the region \([0, \frac{1}{5})\), positive in
\([\frac{1}{5}, \frac{2}{5})\), and the ICE erroneously implies the
existence of heterogeneous effects. As before, UALE quantifies
correctly the ground truth effect, the zero-uncertainty and partitions
the \(x_1\) domain into wide bins that facilitate the interpretation
and create a zero cummulative bin-error \(\mathcal{E}_{\mathcal{Z}}\)
approximation.

@@

\paragraph{(c) With Interaction, Equal weights.}

Here, we activate the interaction term, i.e., \(a=1\), keeping the
weights equal \(a_1=a_2=1\) and \(\sigma_3^2=\frac{1}{4}\). In this
case, it is not straightforward to define the ground-truth effect for
features \(x_1, x_3\), because the interaction term provokes
heterogeneous instance-level effects, i.e., the instance-level effect
of \(x_1\) depend on the unknown value of \(x_3\) and vice-versa. We
observe that UALE's feature effects are quite intuitive. The average
effect of \(x_1\) is the same with the Example (a) with an added
uncertainty, reflecting the uncertainty about the instance-level
effects which is the standard deviation of \(x_3\), i.e.,
\(\sigma_3 = \frac{1}{2}\). The effect of \(x_3\) is only due to the
interaction term, therefore, the instance-level effects
\(\frac{\partial f}{\partial x_3} = x_1 \) follow
\(p(x_1) \sim \mathcal{U}(0,1)\) which has \(\mu_{x_1}=\frac{1}{2}\)
and \(\sigma_{x_1}=\frac{1}{4}\). This is reflected, in UALE's
estimations of Figure~\ref{fig:ex-synth-1-3}.

For \(x_2\), as in Example (a), the effect is
\(f_\mu^{\mathtt{GT}}(x_2)\) is: \(x_2\) in \([0, \frac{1}{4}]\),
\(-x_2\) in \([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in
\([\frac{1}{2}, 1]\)\(x_2\) and it has zero-uncertainty,
\(f_\sigma^{\mathtt{GT}}(x_2) = 0\), since \(x_2\) does not appear in
any interaction term. We confirm that UALE computes it correctly
whereas PDP-ICE fail in both the average effect and the uncertainty.

@@

The above experiments show that UALE models correctly and estimates
accurately both the average effect and its uncertainty. We also
observed that UALE's variable-size bin-splitting method that we
propose, leads to (a) an accurate approximation of the uncertainty and
(b) favoring wider bins we facilitate the interpretation of both the
average effect and the uncertainty. In contrast, PDP-ICE provides
misleading explanations, due to ignoring correlations between
features. The examples above do not cover the case of an interaction
term between correlated features, for example a term \(x_1x_2\),
because in this case there is an open debate about the ground-truth
effect~\citep{Gromping2020MAEP}.
